\section{Outline}
The Python client implements
a simple yet powerful and quite fast
\acronym{api} to interact
with the the \nowdb\ server.
It lives in the module \term{now.py} and provides
the \term{Connection} and the \term{Result} class
as well as a number of handy support functions.

The module uses the package \term{dateutil}
which must be installed on the system.
(Please refer to chapter
\ref{chpt_install} for details.)

\section{Connections}
The \term{Connection} class represents
a \acronym{tcp/ip} connection to the database.
The constructor takes four arguments,
which are all strings:
\begin{itemize}
\item Address:
used to determine the address of the server.
The parameter is passed to the system service
\tech{getaddrinfo} and allows:
an \acronym{ip}v4 address,
an \acronym{ip}v6 address or
a hostname.
Examples: ``127.0.0.1'', ``localhost'',
``myserver.mydomain.org''.

\item Port:
used to determine the port of the server.
The parameter is passed to the system service
\tech{getaddrinfo} and allows:
a port number or a known service name.

\item User:
\comment{currently not used}
\item Password:
\comment{currently not used}
\end{itemize}

The Connection class provides the following 
methods:

\subsubsection{execute(statement)}
The method sends the string \term{statement} to the database
and waits for the response.
It returns an instance of the \term{Result} class
or, on internal errors, raises one of the exceptions
\term{ClientError} or \term{DBError}.

\subsubsection{Variants}
There are two variants of the \term{execute} method,
namely \term{rexecute} and \term{rexecute\_}.
\term{rexecute} suppresses error results.
If the result does not represent an error,
it returns the result just like \term{execute} does.
Otherwise, it releases the result and raises
an exception instead (the leading 'r' stands
for `raise').

\term{rexecute\_} is intended for situations
where the result as such is not interesting,
for instance, when issuing a \acronym{ddl} or
\acronym{dml} statement.
The function always releases the result;
if it is an error,
\term{rexecute\_} would raise an exception,
just like \term{rexecute};
otherwise, it terminates properly,
but does not return anything.

\subsubsection{Single Row Results}
Many \sql\ statements return per definition
only one row. This is true for statements
of the form 

\begin{sqlcode}
\begin{lstlisting}
select count(*), sum(quantity), sum(price) from buys
\end{lstlisting}
\end{sqlcode}

or

\begin{sqlcode}
\begin{lstlisting}
select now()
\end{lstlisting}
\end{sqlcode}

The \term{execute} statement (or one of its variants),
however, would force the programmer to write some
boilerplate code to obtain this single row, \eg:

\begin{python}
\begin{lstlisting}
for row in c.rexecute("select count(*), sum(quantity) from buys"):
    cnt = row.field(0) # see below for details
    sm  = row.field(1)
# use cnt and sm
\end{lstlisting}
\end{python}

For this special use case, there is the method \term{oneRow}:
it receives an \sql\ statement, executes it and handles
the result as a cursor (see below). If it is not a cursor
(or when an error occurs),
the function will raise an exception.
Otherwise, it obtains the first row of this cursor
and appends all the fields in this row to a list.
It releases the cursor and then returns the list.
The above code can thus be simplified to

\begin{python}
\begin{lstlisting}
row = c.oneRow("select count(*), sum(quantity) from buys")
# use row[0] and row[1]
\end{lstlisting}
\end{python}

To go even further: a subset of queries that return only
one row return only one single value. For such such cases,
there is yet another convenience interface, namely \term{oneValue},
which is implemented as

\begin{python}
\begin{lstlisting}
return self.oneRow[0]
\end{lstlisting}
\end{python}

\subsubsection{close()}
The method closes the connection and
releases the C objects allocated by
the constructor. If the connection
cannot be closed for any reason,
the exception \term{ClientError} is raised.

\subsubsection{Resource Manager}
\term{Connection} is a \term{resource manager}.
It is therefore rarely necessary to call
\term{close} directly in user code.
Instead, \term{Connection} can
be used with the \term{with} statement, \ie:

\begin{python}
\begin{lstlisting}
with Connection("localhost", "55505", None, None) as conn:
    # here goes your code
    # refer to the connection as 'conn'
\end{lstlisting}
\end{python}

\section{Results}
The \term{Result} class represents
the dynamic types described in chapter \ref{chpt_sql}.
Instances of \term{Result} are
created and returned by
the \term{Connection.execute()} method.

The \term{Result} class has the following methods:

\subsubsection{rType()}
The method returns the result type,
either \acronym{status}, \acronym{report},
\acronym{row} or \acronym{cursor}.

\subsubsection{ok()}
The method returns \term{True}
if the instance does not represent
an error and \term{False} otherwise.

\subsubsection{release()}
The method releases the C resources
allocated together with the \term{Result} object.
It is not strictly necessary to call
this method in user code, since the
Python \acronym{gc} would invoke it anyway.
However, it is a good policy
to release resources as soon as
they are not needed anymore.
Yet again, \term{Result} is a resource manager
and can be used in \term{with} statements,
which call release internally, when leaving
the block, \eg:

\begin{python}
\begin{lstlisting}
with conn.execute("select count(*) from buys") as cur:
    # here goes your code
    # 'conn' is a previously created connection
    # 'cur' is the result and
    # if no error has occurred, cur is a cursor
\end{lstlisting}
\end{python}

\subsection{Status}
If \term{Result} is a \term{Status},
two more methods are available:

\subsubsection{code()}
The method returns the \nowdb\ error code.
The error code may be
\begin{itemize}
\item 0: Success, no error has occurred;
\item positive:
An error in the database occurred;
chapter \ref{chpt_errors} provides a list
of server-side error codes.
\item negative:
An error in the client library occurred;
chapter \ref{chpt_llc} provides a list
of client-side error codes.
\end{itemize}

\subsubsection{details()}
The method returns
detailed information on the error
(or \term{None} if no error has occurred).

\subsection{Cursors}
If the result is a cursor,
four more methods are available:

\subsubsection{fetch()}
The method fetches the next bulk
of rows from the database.
After successful completion,
the cursor contains this bulk
of rows, which can be obtained by means of
the method \term{row()}.
Note that the first bulk of rows
is available immediately after
the Cursor has been created.

\subsubsection{row()}
The method obtains the current
bulk of rows from the cursor.
It returns a \term{Row} result.

\subsubsection{close()}
The method closes the cursor;
internally, this method is called
wherever, for other results, \term{release}
would be called, that is, by the \acronym{gc}
and when leaving the \term{with} statement
for which the cursor was created.

\subsubsection{eof()}
The method returns \term{True}
if the error state of the cursor
is \term{end-of-file} and \term{False}
otherwise.

\subsubsection{Iterator}
\term{Cursor} is an iterator.
Usually, \term{fetch()}, \term{row()}
and \term{eof()} do not need to be called
explicitly.
Instead a \term{for}-loop  can be used, \eg:

\begin{python}
\begin{lstlisting}
with conn.execute("select * from buys") as cur:
    for row in cur:
        # process row 
\end{lstlisting}
\end{python}

When using a cursor as an iterator
it is automatically released immediately
before the iteration ends.
The above code can therefore be simplified to:

\begin{python}
\begin{lstlisting}
for row in conn.execute("select * from buys"):
    # process row
\end{lstlisting}
\end{python}

\subsection{Rows}
If the result is a row,
two more methods are available:

\subsubsection{field(n)}
The method returns the value
of the $n$th field (starting to count
from 0 for the first field).
The value is an \sql\ base type
converted to Python.
Conversion takes place in the obvious way
(integer and unsigned integer to int,
 float to float, text to string, \etc).
An exception are \term{time} fields.
One could expect \term{field} to return
a \term{datetime}, but that is not the case.
Time is returned as a (signed) integer
representing a \acronym{unix} timestamp
with nanosecond precision.
It can be converted to \term{datetime} with \term{now2dt}
(please refer to section \ref{sec_supp}
for details).

\subsubsection{typedField(n)}
The method is similar to \term{field},
but returns additionally the \nowdb\ type
of the $nth$ field, \eg:

\begin{python}
\begin{lstlisting}
for row in conn.execute("select * from buys"):
    (t,v) = row.typedField(0)
    # t is the type, v is the value
\end{lstlisting}
\end{python}

\subsubsection{count()}
The method returns the number of fields in the row, \eg:

\begin{python}
\begin{lstlisting}
for row in conn.execute("select * from buys"):
    for i in range(row.count()):
        (t,v) = row.typedField(i)
        print("field %d of type %d: %s" % (i, t, v))
\end{lstlisting}
\end{python}

\subsubsection{nextRow()}
The method advances to the next row.
If there was one more row, the method
returns \term{True} and the next call
to \term{field(n)} will return the
$n$th field of that row.
Otherwise, if no more rows are available,
the method returns \term{False}.

Note that, working with a cursor
as iterator, it is not necessary
to use this method. The iterator
already produces single rows.

\subsection{Reports}
If the result is a report,
three more methods are available:

\subsubsection{affected()}
The method returns the number of affected rows.

\subsubsection{errors()}
The method returns the number of errors.
Note that \acronym{dml} statements will
return a \term{Status} result if an error occurred.
In practice, only \acronym{dll} statements provide
this information.

\subsubsection{runTime()}
The method returns the running time of 
the \acronym{dml} or \acronym{dll} statement.

A usage example is

\begin{python}
\begin{lstlisting}
with conn.execute("load '/opt/import/client.csv into client") as rep:
    if not rep.ok():
        print("ERROR: %s " % rep.details())
        return
    print("affected: %d, errors: %d, running time: %dus" % 
             (rep.affected(), rep.errors(), rep.runTime()))
\end{lstlisting}
\end{python}

\section{Exceptions and Errors}
\subsubsection{ClientError}
Is raised when an error in the client library occurred
(\eg\ a socket error).

\subsubsection{DBError}
Is raised when an error in the database occurred.

\subsubsection{WrongType}
Is raised in case of type mismatch, \ie\
trying to call a method not available for
this specific return type.

\subsubsection{explain(err)}
The function \term{explain} returns
a description of the error code passed in.

\section{Support Functions}\label{sec_supp}
\subsubsection{dt2now(dt)}
The function expects a \term{datetime}
object and converts it to a \nowdb\ timestamp.

\subsubsection{now2dt(tp)}
The function expects a \nowdb\ timestamp
and converts it to a \term{datetime} object.

\subsubsection{utc}
This is a global variable defined in \term{now.py}.
It is the \acronym{utc} timezone object needed
for many datetime constructors and conversion functions.
It is provided for convenience.

\subsubsection{TIMEFORMAT and DATEFORMAT}
These are global constants providing the
standard \nowdb\ time and date format
strings that can be used, for instance,
in the \term{datetime}
\term{strftime} and \term{strptime} methods, \eg:

\begin{python}
\begin{lstlisting}
    stmt = "select count(*) from buys \
             where timestamp = '" + tp.strftime(TIMEFORMAT) + "'"
\end{lstlisting}
\end{python}
